Why does female meiosis allow the opportunity for exploitation by selfish sperm? BETTER TITLE?

### 2013-09-14 writeup_and_notes/male_dependent_drive.tex
Yaniv's changes upto page 5 on Friday

%Since all alleles within an individual rely on individual survival and reproduction for evolutionary success, most of life in a diploid eukaryotic genome is harmonious.
%However, this harmony is incomplete, as in numerous cases an allele can benefit (in the short or long term) from taking advantage of its host.
Despite the apparent unity of the organism, occasionally alleles can
gain an evolutionary advantage at a cost to  individual fitness
\citep{Burt2006}, often by exploiting meiosis and gametogensis.
%A clear opportunity for this conflict arises during gametogenesis where alternative alleles are in direct competition for representation in the next generation.
Female meiosis, an asymmetric event in which only one of two alternate alleles
enters the  egg, with the others consigned to the polar body, is one such occasion \citep{Sandler1957,Pardo-ManuelDeVillena2001a}. 
%Female meiosis is particularly unfair, only one of the four products of meiosis becomes an egg, while the other threes are disgarded into polar bodies. 
An allele that biases female meiosis in its favor (i.e. a meiotic driver), may increase in frequency even if this driver entails a pleiotropic fitness cost \citep{Prout1973}, generating a genetic conflict between the success of the driver and organismal fitness.
%Alleles that can segregate into the egg more than half the time in heterozygotes, by exploiting asymmetries, 
%can potentially increase in frequency in the population (i.e. experiencing true meiotic drive).
%If such alleles have fitness consequences they are a source of genetic conflict, 
%and can become balanced in the population if their host in homozygotes 
%out weighs their ability to spread to spread through heterozygotes. 
Meiotic drivers observed in nature (in both plants
\citep{Buckler1999,Fishman2005,Fishman2008}, and animals
\citep{Agulnik1990,Wu2005,Pardo-ManuelDeVillena2001c}) highlight this
conflict -- the selfish benefits of drive and the associated
pleiotropic fitness costs sustain a balanced polymorphism
\citep{Prout1973}, 
and often generate on ongoing evolutionary escalation of drive suppressors and enhancers \citep{Dawe1996,Fishman2008}. 
%Such balanced drive systems can cause the subsequent evolution of linked enhancers of drive
%and of supressors of drive throughout the genome. 
%Indeed, the known polymorphic female meiotic drive systems (OM, maize knobs, IN, mimulus, others?) habor
%a great diversity of enhancers and supressors. 
The threat of meiotic drive to organismal fitness is potentially so
severe that it has been hypothesized that many basic properties of
meiosis and o\"{o}genesis, including the initial genome doubling in
meiosis I \citep{Haig1991}, arrested female meiosis
\citep{Mira1998}, centromere mechinery \citep{MalikHenikoff}, and sex differences in the recombination rate \citep{Haig2010,Brandvain2012} have evolved to disrupt meiotic drive and enforce fairness.

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, the completion of female meiosis requires fertilization of the egg, 
and there is ample opportunity for interaction between the sperm and female meiotic machinery.
If, for example, an allele in sperm could facilitate meiotic drive by a genetically equivalent allele in a 
heteromorphic dyad, such an allele could presumably bias meiosis in its favor and rapidly spread through the population.
At first sight, it seems as although female meiosis is primed to be exploited by selfish sperm systems.

@@

Why then is the requirement of fertilization to complete female meiosis so ubiquitous? 
It is certainly not the case that animals are mechanistically incapable of evolving past this requirement.
There is considerable variation in which stage of meiosis requires fertilization, and 
a number of animal clades (should we try and lower bound how many transitions?) have evolved
to allow the completion of female meiosis upon ovulation. 
%(Although even in these organisms female meiosis 
%may also be subjection to influence by sperm alleles, if the sperm arrives at the egg soon enough after 
%ovulation to influence meiosis.)

It is also not the case that sperm is mechanistically incapable of influencing the outcome of female meiosis.
%This is not just idle speculation, as there is also direct evidence that sperm components and indeed different sperm genotypes can affect female meiosis.
%are mechanistic reasons to think that this is more than idle speculation.
Mechanistic evidence for this possibility comes from
\emph{C. elegans}, where experimentally suppressing XXX leads to
premature deployment of the aster (a vital component of mitotic
machinery) provided by the sperm, disrupting MII meiotic segregation in the egg, leading to a triploid zygote. 
%This demonstrates that physical components derived from the sperm are are capable of affecting the outcome of female meiosis.
Additionally, genetic evidence suggests that the transmission patterns
in heterozygous females many potentially depend on sperm haplotype. 
Specifically, the two best characterized female meiotic drive systems in mouse (In and Om), both operate by distorting the second meiotic division, 
and in both systems the outcome of female meiosis depends the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.

In this article we explore through simple population genetic models the consequences of alleles that influence the outcome 
of female meiosis. 
We use these models to argue that it is actually surprisingly hard for the influence of
sperm on the outcome of meiosis to drive sustained conflict. In fact we find that sperm and egg 
genomes' interests are often aligned as they are both invested in the fate of the zygote they will form (as was suggested for the In locus \citep{Pomiankowski1993}).
This suggests that females are unlikely to evolve to prevent the influence of sperm on meiosis,
and indeed features of meiosis may evolve that facilitate the interaction of sperm with female meiosis.

@@

The standard model of female meiotic drive is two allele model, where the
allele $1$ is transmitted $d$ fraction of the time through female
meiosis ($d > \frac{1}{2}$), regardless of the genotype of the sperm. 
In mammals, fertilization takes place at MII, so we imagine this drive must be taking place at MII in order
for a sperm to have any influence. For drive to take place at MII there has to be an uneven number of
 crossovers between the centromere and the drive locus, such that realistically $d$ is bounded to be $<0.75$?. 
In other systems where fertilization occurs during MI, sperm could influence either drive at MI or MII, 
and drivers at MI can have a $d=1$ if they occur in tight linkage with the centromere.

We can model a simple sperm dependent meiotic drive system, by
modifying this model so that the allele only drives during female
meiosis if it fertilized by a sperm carrying that allele
 (see Figure \ref{Eggsperm_2_allele_cartoon}). Under what conditions can
this self propograting driver allele spread, and could it drive the
evolution of supressors?

We first consider the situation where the driving allele has no
fitness consequences in heterozygotes, only in homozygotes (whose
relative fitness is $1-s$). A standard meiotic drive allele can always
invade the population in this situation, as it is initially only
present mostly in heterozygotes and rarely suffers the consequences of
the problems in homozygotes. However, if the cost in homozygotes is
strong enough ($s>XXd$), then the allele is prevented from reaching
fixation and instead is maintained as a balanced polymorphism in the
population \citep{Prout1973}. Such balanced polymorphisms can then
drive the evolution of supressors of drive \citep{XX}, and most of our
examples of drive systems are of this form.

Under a recessive fitness cost model our sperm dependent driver
actually has far more restrictive conditions to invade the population,
than a standard meiotic driver (see Figure \ref{Invasion_space}).
That is because our allele which drives only when it is present in the
fertilizing sperm is guarenteed to form a homzygote. Thus, 
to even spread in the population when it is rare, our sperm-dependent
driver has to overcome the cost it suffers in the homozygote state.

When conditions are suitable for such alleles to spread into the
population, they spread slowly through low frequencies as they rarely
drive as they are unlikely to be fertilized by a sperm with the right
allele. Once they become established they spread rapidly to fixation due to their
positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}). 
Unlike standard meiotic drive systems 
these alleles can not become balanced in the population  by homozygote
cost, as they have already over come this cost just to enter the population.

Things are even worse for the sperm-dependent driver if it suffers any
fitness cost in the heterozygote. Our standard driver can invade the
population from low frequency if its drive in heterozygotes out weighs
the fitness cost they suffer (XXX). However, when at low frequency our
sperm dependent driver is often in heterozygotes but will only drive
if fertilized by the same allele. If our allele incurs even very weak costs
in heterozygotes it can be prevented from spreading from very low
frequencies, unless its drive is very strong. The dynamics of 
our sperm-dependent drivers with an additive cost resemble an underdominant polymorphism
(in that the system is bistable). Below a cutoff frequency the allele is
unable to enter the population (in a deterministic model) as it is
paying the cost of the drive but drive is ineffective (see Figure
S\ref{bistable_additive}). Above this
frequency the allele increases deterministically as its drive is sufficiently effective 
becayse heterozygote eggs are fertilized by sperm with our allele
sufficiently often to beat the costs. 
This bistability suggests that sperm-dependent drivers will often be 
unable to enter into reasonably sized populations, and if they do they
should rapidly be spread fixation.

>>>>

%Since all alleles within an individual rely on individual survival and reproduction for evolutionary success, most of life in a diploid eukaryotic genome is harmonious.
%However, this harmony is incomplete, as in numerous cases an allele can benefit (in the short or long term) from taking advantage of its host.
Despite the apparent unity of the organism, occasionally alleles can
gain an evolutionary advantage at a cost to  individual fitness
\citep{Burt2006}, often by exploiting meiosis and gametogensis.
%A clear opportunity for this conflict arises during gametogenesis where alternative alleles are in direct competition for representation in the next generation.
Female meiosis, an asymmetric event in which only one of two alternate alleles
enters the  egg, with the others consigned to the polar body, \yb{particularly vulnerable to such exploitation}  \citep{Sandler1957,Pardo-ManuelDeVillena2001a}. 
%Female meiosis is particularly unfair, only one of the four products of meiosis becomes an egg, while the other threes are disgarded into polar bodies. 
An allele that biases female meiosis in its favor (i.e. a meiotic driver), may increase in frequency even if this driver entails a pleiotropic fitness cost \citep{Prout1973}, generating a genetic conflict between the success of the driver and organismal fitness.
%Alleles that can segregate into the egg more than half the time in heterozygotes, by exploiting asymmetries, 
%can potentially increase in frequency in the population (i.e. experiencing true meiotic drive).
%If such alleles have fitness consequences they are a source of genetic conflict, 
%and can become balanced in the population if their host in homozygotes 
%out weighs their ability to spread to spread through heterozygotes. 
Meiotic drivers observed in nature (in both plants
\citep{Buckler1999,Fishman2005,Fishman2008}, and animals
\citep{Agulnik1990,Wu2005,Pardo-ManuelDeVillena2001c}) highlight this
conflict -- the selfish benefits of drive and the associated
pleiotropic fitness costs sustain a balanced polymorphism
\citep{Prout1973}, 
and often generate on ongoing evolutionary escalation of drive suppressors and enhancers \citep{Dawe1996,Fishman2008}. 
%Such balanced drive systems can cause the subsequent evolution of linked enhancers of drive
%and of supressors of drive throughout the genome. 
%Indeed, the known polymorphic female meiotic drive systems (OM, maize knobs, IN, mimulus, others?) habor
%a great diversity of enhancers and supressors. 
The threat of meiotic drive to organismal fitness is potentially so
severe that it has been hypothesized that many basic properties of
meiosis and o\"{o}genesis, including the initial genome doubling in
meiosis I \citep{Haig1991}, arrested female meiosis
\citep{Mira1998}, centromere machinery \citep{MalikHenikoff}, and sex differences in the recombination rate \citep{Haig2010,Brandvain2012} have evolved to disrupt meiotic drive and enforce fairness.

It is therefore somewhat surprising that despite the intense evolutionary pressure on female meiosis to prevent meiotic drive, 
it is potentially open to sabotage by a virtual stranger -- a haploid sperm genome.
That is, in many animal species, the completion of female meiosis requires fertilization of the egg, 
and there is ample opportunity for interaction between the sperm and female meiotic machinery.
If, for example, an allele in sperm could facilitate meiotic drive by a genetically equivalent allele in a 
heteromorphic dyad, such an allele could presumably bias meiosis in its favor and rapidly spread through the population.
At first sight, it seems as although female meiosis is primed to be exploited by selfish sperm systems.  
\yb{Why then is the requirement of fertilization to complete female meiosis so ubiquitous? }

@@

%Why then is the requirement of fertilization to complete female meiosis so ubiquitous? 
It is certainly not the case that animals are mechanistically incapable of evolving past this requirement.
There is considerable variation in which stage of meiosis requires fertilization, and 
a number of animal clades (should we try and lower bound how many transitions?) have evolved
to allow the completion of female meiosis upon ovulation. 
%(Although even in these organisms female meiosis 
%may also be subjection to influence by sperm alleles, if the sperm arrives at the egg soon enough after 
%ovulation to influence meiosis.)

It is also not the case that sperm is mechanistically incapable of influencing the outcome of female meiosis.
%This is not just idle speculation, as there is also direct evidence that sperm components and indeed different sperm genotypes can affect female meiosis.
%are mechanistic reasons to think that this is more than idle speculation.
Mechanistic evidence for this possibility comes from
\emph{C. elegans}, where experimentally suppressing XXX leads to
premature deployment of the aster (a vital component of mitotic
machinery) provided by the sperm, disrupting MII meiotic segregation in the egg, leading to a triploid zygote. 
%This demonstrates that physical components derived from the sperm are are capable of affecting the outcome of female meiosis.
Additionally, genetic evidence suggests that the transmission patterns
in heterozygous females many potentially depend on sperm haplotype. 
Specifically, the two best characterized female meiotic drive systems in mouse (In and Om), both operate by distorting the second meiotic division, 
and in both systems the outcome of female meiosis depends the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}; 
	\yb{however,  we acknowledge the difficulty in reeling out early mortality as an alternative explanation of these putative drivers}.

Here, we develop population genetic models exploring the evolution of alleles in sperm that influence the outcome of female meiosis 
%In this article we explore through simple population genetic models the consequences of alleles that influence the outcome of female meiosis.
These models argue that it is actually surprisingly hard for the influence of sperm on the outcome of meiosis to drive sustained conflict. 
In fact, the interests of sperm and egg  genomes' are often aligned, as are both invested in the fate of the resultant zygote 
	(as was speculated for the In locus \citep{Pomiankowski1993}).
Therefore, there is little selective benefit to females in preventing sperm to influence female meioses,
	indeed features of meiosis may evolve that facilitate the interaction of sperm with female meiosis.

@@

In the standard model of female meiotic drive, the driving allele at a balletic site is transmitted to the egg with probability  $d > \frac{1}{2}$, regardless of sperm genotype. 
%The standard model of female meiotic drive is two allele model, where the allele $1$ is transmitted $d$ fraction of the time through female meiosis ($d > \frac{1}{2}$), regardless of the genotype of the sperm. 
\yb{ To depict a case of a self-promoting (or `green-bearded') meiotic driver,  we modify this standard model such that the female meiotic driver is only effective when fertilized by a sperm carrying that allele (see Figure \ref{Eggsperm_2_allele_cartoon}).}
We then identify the conditions under which this self propagating driver can spread, and if it drive the evolution of suppressors? 
\yb{We note that the timing of fertilization relative to female meiosis places another constrain ton $d$, 
	for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
	so $d$ is bounded to be $<0.75$?.}  
\yb{I THINK THIS ORDER IS A BIT BETTER, BUT STILL ``OFF'', perhaps this is a footnote, or in methods or something...}
%In mammals, fertilization takes place at MII, so we imagine this drive must be taking place at MII in order
%for a sperm to have any influence. For drive to take place at MII there has to be an uneven number of
% crossovers between the centromere and the drive locus, such that realistically $d$ is bounded to be $<0.75$?. 
%In other systems where fertilization occurs during MI, sperm could influence either drive at MI or MII, 
%and drivers at MI can have a $d=1$ if they occur in tight linkage with the centromere.

%We can model a simple sperm dependent meiotic drive system, by
%modifying this model so that the allele only drives during female
%meiosis if it fertilized by a sperm carrying that allele
 %(see Figure \ref{Eggsperm_2_allele_cartoon}). Under what conditions can
%this self propagating driver allele spread, and could it drive the
%evolution of supressors?

\yb{We first consider the situation where the driving allele is neutral in heterozygotes, 
	but deleterious in homozygotes (whose relative fitness is $1-s$). } 
%We first consider the situation where the driving allele is has no
%fitness consequences in heterozygotes, only in homozygotes (whose
%relative fitness is $1-s$). 
\yb{In this model, a standard meiotic driver can always invade because 
	when rare it occurs predominantly in heterozygotes and therefore it initially pays no fitness cost.} 
%A standard meiotic drive allele can always
%invade the population in this situation, as it is initially only
%present mostly in heterozygotes and rarely suffers the consequences of
%the problems in homozygotes. 
\yb{However, a severe homozygotes fitness cost ($s>XXd$) prevents  
	drivers from fixing  \citep{Prout1973}, 
	and provides an opportunity for the
	evolution of drive suppressors  \citep{XX}, corresponding well to 
	empirical examples of female meiotic drive. }

%However, if the cost in homozygotes is
%strong enough ($s>XXd$), then the allele is prevented from reaching
%fixation and instead is maintained as a balanced polymorphism in the
%population \citep{Prout1973}. Such balanced polymorphisms can then
%drive the evolution of supressors of drive \citep{XX}, and most of our
%examples of drive systems are of this form.

\yb{In contrast to a traditional driver, which can drive but pay no fitness cost when rare, 
	our green-bearded driver drives by creating homozygotes, and therefore pays a fitness cost even when rare. 
To simply increase when rare, our sperm-dependent driver must overcome a homozygous fitness cost.
Therefore with a recessive fitness cost,
	the conditions allowing the invasion of our green-bearded driver 
 	are far more restrictive than those for a standard meiotic driver (see Figure \ref{Invasion_space}).}

%Under a recessive fitness cost model our sperm dependent driver
%actually has far more restrictive conditions to invade the population,
%than a standard meiotic driver (see Figure \ref{Invasion_space}).
%That is because our allele which drives only when it is present in the
%fertilizing sperm is guarenteed to form a homzygote. Thus, 
%to even spread in the population when it is rare, our sperm-dependent
%driver has to overcome the cost it suffers in the homozygote state.

\yb{When a deleterious recessive green-bearded driver does spread, 
	it spends much time at low frequency, because when rare it is unlikely to meet a complimentary sperm. 
However, once common, a green bearded driver spreads rapidly to fixation due to their
	positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}).  PERHAPS COMPARE TO TRADITIONAL DRIVER IN THIS FIGURE
Unlike standard meiotic drive systems, polymorphism of a green-bearded driver cannot be sustained by a homozygous cost, 
	because it already overcame this cost just to enter the population.
The lack of a balanced polymorphism precludes the evolution of an allele that suppresses this form of meiotic drive.}

%When conditions are suitable for such alleles to spread into the
%population, they spread slowly through low frequencies as they rarely
%drive as they are unlikely to be fertilized by a sperm with the right
%allele. Once they become established they spread rapidly to fixation due to their
%positive frequency dependent behavior (Inset in Figure \ref{Invasion_space}). 
%Unlike standard meiotic drive systems 
%these alleles can not become balanced in the population  by homozygote
%cost, as they have already over come this cost just to enter the population.

\yb{Inclusion of a heterozygous fitness cost (e.g. additivity) further constrains the evolution of a sperm-dependent driver. 
A  standard female  driver can invade from low frequency so long as drive in heterozygotes out weighs the fitness cost they suffer (XXX); 
	however, when at low frequency our sperm dependent driver pays the heterozygous fitness cost but rarely drives because it is rare in sperm.
Therefore, even very weak heterozygous fitness costs
	prevent the spread of a very rare green-bearded driver, unless its drive is very strong. 
With an additive cost, the dynamics of a green-bearded driver resembles that of an underdominant mutation (in that the system is bistable). 
Below a cutoff frequency the allele deterministically decreases in frequency 
	because it pays the  heterozygous fitness cost  but drives is ineffectively (see FigureS\ref{bistable_additive}). 
Above this frequency, the allele increases deterministically because the commonness of sperm carrying this allele 
	make drive effective enough to overcome the associated fitness cost.
%becayse heterozygote eggs are fertilized by sperm with our allele %sufficiently often to beat the costs. 
This bistability will  prevent sperm-dependent drivers from invading 	
	reasonably sized populations, and assures that if they do invade they will rapidly fix.}

%Things are even worse for the sperm-dependent driver if it suffers any
%fitness cost in the heterozygote. Our standard driver can invade the
%population from low frequency if its drive in heterozygotes out weighs
%the fitness cost they suffer (XXX). However, when at low frequency our
%sperm dependent driver is often in heterozygotes but will only drive
%if fertilized by the same allele. If our allele incurs even very weak costs
%in heterozygotes it can be prevented from spreading from very low
%frequencies, unless its drive is very strong. The dynamics of 
%our sperm-dependent drivers with an additive cost resemble an underdominant polymorphism
%(in that the system is bistable). Below a cutoff frequency the allele is
%unable to enter the population (in a deterministic model) as it is
%paying the cost of the drive but drive is ineffective (see Figure
%S\ref{bistable_additive}). Above this
%frequency the allele increases deterministically as its drive is sufficiently effective 
%becayse heterozygote eggs are fertilized by sperm with our allele
%sufficiently often to beat the costs. 
%This bistability suggests that sperm-dependent drivers will often be 
%unable to enter into reasonably sized populations, and if they do they
%should rapidly be spread fixation.

XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX
EDITED TO HERE
XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX
